The Countercyclical Component of Capital Regulation

### 2020-05-10 Analysis/Draft/Draft_bookdown.Rmd
Several changes to the draft text. Version before I started the RW-lending project

---
title: "The Countercyclical Component of Capital Regulation"
date: '\emph{\today}'
output:
  bookdown::pdf_document2:
    fig_caption: yes
    latex_engine: pdflatex
    template: tex-ms.tex
  pdf_document: default
  word_document: default
author:
- name: "Lucas Avezum"
  affiliation: "CentER and EBC, Tilburg University"
biblio-style: apsr
bibliography: ../../Auxiliar/mybibfile.bib
citecolor: black
csl: ../../Auxiliar/cambridge-journal-of-economics.csl
#anonymous: no
fontsize: 11pt
geometry: margin= 1in
header-includes: \usepackage{rotating, graphicx, dcolumn, longtable, float, siunitx, tabularx}
keywords: Internal ratings-based models, Risk-weighted assets dispersion, Procyclicality
linkcolor: black
abstract: The Basel II accord introduced a concave mapping from probabilities of default
  (PD) to capital requirements for banks adopting the internal ratings-based (IRB)
  approach. Therefore, total capital requirements depends not only on the portfolio's
  average PD but also on the shape of its PD distribution. Using a hand-collected
  dataset, I propose a method that decomposes the contribution of the PD distribution
  to capital requirements into an average component and a shape component. While a
  larger average component increases capital charges, a larger shape component generates
  capital savings. I investigate the relationship of these capital savings with the business cycle and find it to   be countercyclical. I further document that the shape are important sources of across banks. The results suggest   that a recalibration of the current IRB framework can have sizeble effects on the procyclicality of capital       regulation.
spacing: double
subtitle: ''
thanks: Preliminary draft.
JEL: G21, G28, G11
urlcolor: black
toc: FALSE
---

@@

library(tidyverse)            ## Data manipulation, pipe operator
library(data.table)           ## Command to bind lists
library(ggridges)             ## Density ridges plot
library(stargazer)            ## Latex tables
library(lfe)                  ## High-dimensional fixed effects
library(gridExtra)            ## Create grid for plots and tables
library(cowplot)              ## Create grid for plots and tables
library(DescTools)            ## Command to winsorize variables
library(ggpubr)               ## More plots
library(splitstackshape)     ## Command to expand data

#library(textablr)
#library(sandwich)
#library(lmtest)
#library(relaimpo)            ## Calculate relative importance measures

@@

# Introduction

The regulation on bank capital requirements have evolved substantially since the first international standard, the 1988 Basel Accord. At the heart of these changes is the link of capital charges to asset risk [@basel2004international; @BCBS2017basel; @EBA2019policy]. Risk sensitivity of capital requirements was present in the first accord and it was greatly enhanced in the Basel II framework by giving banks the option to calculate capital requirements using risk estimates obtained from their own models, known as the internal ratings-based (IRB) approach. The rationale for risk-linked capital charges is that regulating capital via a simple capital to asset ratio can incentivize banks to take on excessive risk [@koehn1980regulation; @kim1988risk; @gordy2010risk]. The rationale for allowing internal models is the better alignment of risk estimates to actual portfolio risk [@barakova2014banks] and the promotion of better risk management practices [@cucinelli2018credit].

However, both risk sensitivity and the use of internal models may have adverse consequences. First, linking capital charges to asset risks exacerbate the procyclicality of lending [@danielsson2001academic; @kashyap2004cyclical; @repullo2012procyclical; @behn2016procyclical]. As the overall credit quality detiorates during economic downturns, capital requirements tighten, forcing banks to cut down lending. Second, the inherent flexibility of the IRB approach may allow for differences in capital requirements that do not reflect portfolio risk but rather modelling choices [@le2012revisiting; @mariathasan2014manipulation; @behn2016limits;@berg2017analysis; @ferribank].

@@

With this novel dataset, I start by documenting the relative importance of *Capital savings* to banks' overall RW. By extending the decomposition method proposed by @cannata2012inside, I find that *Capital savings* accounts for 27% of the RW variation across banks. I also find that *Capital savings* measures are negatively correlated to other RW components. This correlation implies that banks with higher average portfolio riskness benefit the most from *Capital savings*. Lastly, although banks benefited from higher *Capital savings* as they moved their portfolio from the standardized to the IRB approach (known as the roll-out effect), the intensity of *Capital savings* has, on average, decreased in time. That is, for the same average PD, capital requirements have increased from 2008 to 2018.

@@

I further analyse the relationship of the PD distribution moments with GDP growth and find that during economic expasions average PD and PD variance decrease while the PD kurtosis and PD skewness increase. This evidence implies that distributions become more symmetric and with less extreme values during recessions, suggesting a flight to safety behaviour by banks. As the recession reduces their existing portfolio credit quality (evidenced by the negative coefficient of average PD) banks respond by reducing risk-taking.

@@

The remainder of this paper is organized as follows. Section 2 describes the capital requirement formula, the *Capital savings* measure, and the hypothesis concerning this measure. Section 3 describes the collected data and analyses the relative importance of *Capital savings* in explaining differences in risk-weights across banks and time. Section 4 presents evidence on the countercyclical feature of *Capital savings*. Finally, section 5 concludes.

# Institutional Background and Capital Savings Measures

## Capital Requirements under the IRB approach

In June 2004, the Basel Committee issued a revised framework on international convergence of capital measurements and capital standards [@basel2004international], which serves as the basis for national rulemaking and implementation processes. Since then, financial institutions may choose between two approaches to calculate the capital requirement for credit risk; the standardized approach (essentially a slightly modified version of the first accord) and the IRB approach.

In the IRB approach, capital requirements are derived from risk-weight formulas that take as input banks' own estimates of credit risk parameters. The basic regulatory formula derives from a model characterized by the property that the risk-weight of each asset depends only on the estimated credit risk parameters for this asset and not on the composition of the portfolio, i.e. the model is portfolio invariant. This feature leads to a bottom-up approach where capital requirements are determined on the asset level and total requirement is simply the sum of asset individual requirement.

Equation \ref{eq1} is the basic risk-weight formula:

\begin{equation}
\label{eq1}
\text{RW}= 12.5\frac{\sum_{i}\text{EAD}_i\text{LGD}_i N\bigg(\sqrt{\frac{1}{1-R}}N^{-1}(\text{PD}_i)+\sqrt{\frac{R}{1-R}}N^{-1}(0.999)\bigg)}{\sum_{i}\text{EAD}_i} ,
\end{equation}

where $i$ is an asset in the portfolio and $N$ is the standard normal cumulative distribution. Equation \ref{eq1} calculates the average RW for the porftolio. The total amount of RWA for this portfolio is $\text{RW}\times \sum_i\text{EAD}_i$ and total capital requirements $\text{RWA}\times 8\%$.

Basel accords segment credit portfolio into several categories, each having different additional features (maturity adjustments, risk-dependent correlation coefficient, among others) while keeping the basic structure of equation \ref{eq1}. The credit risk parameters estimated by banks, which are used as input on equation \ref{eq1}, are: Exposure at default (EAD), loss given default (LGD), and probability of default (PD).

## Probability of Default Distribution and Capital Savings

Figure \ref{fig:pdmap} plots the mapping from PD to RW according to equation \ref{eq1}. For my analysis, the crucial characteristic is that this mapping is concave. The concavity of the PD-RW mapping together with the bottom-up method to aggregate capital requirements imply that actual capital requirements are lower than if they were calculated using the average PD to the entire portfolio. For clarity, consider two portfolios. The first contains a low PD asset and a high PD asset in equal proportions. The second has the same total amount as the first but contains only a mean PD asset, where mean PD is the mean of low PD and high PD. Because of the concavity of the IRB formula, by Jensen’s inequality, the capital requirement for the second portfolio is greater than the first. More generally, total capital requirements depend not only on banks' portfolio average PD but also on the shape of the PD distribution. Figure \ref{fig:pdmap} illustrates the comparison. $\text{RW}^a$ and $\text{RW}^c$ are the average RW for the first portfolio and the second portfolio, respectively. The difference between  $\text{RW}^c$ and $\text{RW}^a$ is the monetary amount of *Capital savings* per unit of EAD due to the concavity of the IRB formula.

I propose a method that decomposes the contribution of the PD distribution to capital requirements into an average component and a shape component. The method is a generalization of the example in Figure \ref{fig:pdmap}: for each bank-year combination, compare the actual capital requirement to the counterfactual case where the requirement for the entire portfolio is calculated using the portfolio’s average PD.

I define the counterfactual RW as:

\begin{equation}
\label{eq2}
\text{RW}^c= 12.5\frac{\sum_{i}\text{EAD}_i\text{LGD}_iN\bigg(\sqrt{\frac{1}{1-R}}N^{-1}(\overline{\text{PD}})+\sqrt{\frac{R}{1-R}}N^{-1}(0.999)\bigg)}{\sum_i \text{EAD}_i},
\end{equation}
in which the PD parameter for all assets is replaced by the portfolio weighted average:

@@

The main measure of *Capital savings* used in this paper is the ratio of equation \ref{eq2} to equation \ref{eq1} as follows:

@@

$\text{RW}^r$ is the rate at which a bank saves capital. For instance, a value of 0.5 means this portfolio would require 50% more capital in the abscense of *Capital savings* to keep the same capital ratio. A value of zero means that there are no *Capital savings* and implies a portfolio with a degenerate PD distribution. Any PD distribution with positive variance results in positive *Capital savings*.

@@

\begin{equation}
\label{eq:measure3}
\text{RW}^s=\frac{\text{RWA}^{\$}}{\text{RWA}}.
\end{equation}

# Data

@@

Figure \ref{fig:savingsbank} shows that the *Capital savings ratio* calculated with the compiled dataset. Each point is a bank-year observations and the bars indicates the average value across time for each bank. There is substancial variation across banks with Goldman Sachs saving the highest at an average rate of 120% and Mediobanca the lowest at an average rate of 18.4%. Variation across time is considerable with, for instance, *Capital savings ratio* for Deutsche Bank going from 108.7% in 2010 to 61.4% in 2014. *Capital savings ratio* also varies between and within countries. Figure \ref{fig:savingscountry}  aggregates the dataset by country. In the figure, each point is a bank observation and the bars are the average of these points. Variation between countries reaches its maximum at 50% (Denmark versus Norway) and within at 75% (between BNY Mellon and Goldman Sachs in the United States).

# Capital Savings and the Business Cycle

In this section I investigate the relationship between *Capital savings* and the business cycle. I start by explaining the theoretical ambiguous relationship between *Capital savings* and the business cycle. Next I show that *Capital savings* are countercyclical, i.e., given an average portfolio PD, banks need more capital in good times than in bad times to reach the same capital ratio. Finally, I provide evidence that this relationship is not purely mechanical but depends on how banks realocate their assets through the business cycle.

## A simple theoretical framework {#theory}

Overall, capital requirements are procyclical: during economic downturns credit risk in the economy increases which raises capital requirements, under the IRB approach forcing banks to cut down lending. However, as credit conditions changes (and banks react to it) the amount of *Capital savings* also changes. The expected direction of *Capital savings* cyclicality is ambiguous. On the one hand, if the entire portfolio is equally affected, capital savings are procyclical. On the other hand, if riskier assets are more sensitive to credit shocks, then capital savings are countercyclical. The ambiguity arises because *Capital savings* can only increase, when credit conditions deteriorates, if the credit distribution widens. In the first situation, there is only a shift of the location parameters of the distribution and, consequently, a reduction of *Capital savings*.

I then extend the analysis to investigate the relationship between *Capital savings* and the business cycle. Prior evidence suggests that capital requirements are procyclical: during economic downturns banks tend to deleverage instead of raising capital. However, as credit conditions deteriorates (and banks react to these new credit conditions) the PD distribution and, consequently, the amount of *Capital savings* also change. A priori the direction of the economic cycle effect on *Capital savings* is ambiguous. At one extreme, if the entire portfolio is equally affected, then there is only a shift of the location parameter of the PD distribution. Because the PD-RW formula elasticity is lower for higher PD estimates, in this scenario *Capital savings* are procyclical. On the other hand, if riskier assets are more sensitive to credit shocks, then capital savings are countercyclical. The ambiguity arises because *Capital savings* can only increase during a recession if the shape of the PD distribution is affect during the cycle.

I use a simple example to illustrate this ambiguity. Consider a bank with the asset side of its balance sheet composed of \$100 of a safer asset and \$100 of a riskier asset and three states of the economy. Without loss of generality, assume $\text{LGD}=1$ and $\text{R}=0.15$. In the initial state, safer and riskier assets have a 0.1% and 0.2% probability of default, respectively. According to equation \ref{eq1}, the actual RWA of this bank is \$63.9. The weighted average of this portfolio in terms of $\text{PD}$ is 0.15% which yields a counterfactual RWA of \$64.7 according to equation \ref{eq2}. Hence, in the initial state *Capital savings* equal \$0.8.

Consider now a credit risk shock. In state A, both safer and riskier assets are affected, having their PDs raised to 0.2% and 0.3%, respectively. In state B, only the riskier asset is affected, having its PD equal to 0.4%. Note that the average PD in both states are equal at 0.25%. However, relative to the initial state, the capital savings measure for state A decreases to \$0.45 while it increases to \$4.35 in state B. The mechanism of this example is simple: relative to the initial state, in state A only the location parameter of the PD distribution changes (standard deviation is equal to 0.05% in both states) while in state B also other moments change (standard deviation increases to 0.15%). Note that in both state A and B total capital requirements increase as average credit risk is higher relative to the initial state in both states. This net procyclical effect is in line with the results of @behn2016procyclical. However, due to the increased portfolio dispersion, capital savings are countercyclical in state B. Conversely, because the PD-RW formula elasticity is lower for higher PD estimates, in stata A *Capital savings* are procyclical. Therefore, capital savings are procyclical in state A. Table \ref{tab1} summarizes the example and Figure \ref{fig:pdmapcases}

Whether *Capital savings* are pro- or countercyclical is an empirical question, as the direction of the cyclicality depends on how the economic cycle affects the PD distribution.

## Empirical Results

@@

The dependent variable is the yearly change in a variable associated to portfolio $j$ for bank $i$ at year $t$. The change in the logatithm of real GDP per capita used comes from the country where the bank has its headquarter. All regressions include bank and year fixed effects and robust standard errors adjusted for clustering at the country level are reported in parentheses.

Column 1 Table \ref{tab:main} shows results using equation \ref{eq:measure1}, my prefered measure of *Capital savings*.  shows that total *Capital savings* are countercyclical. According to regression 1 an increase of 1.6 percentage points in GDP growth rate (one standard deviation) decreases total *Capital savings* growth rate by 0.7 percentage points on average. The effect is economically sizeable representing 14% of total *Capital savings* standard deviation in the sample.

@@

## Economic Impact of Capital Savings Cyclicality

>>>>

---
title: "The Countercyclical Component of Capital Regulation"
date: '\emph{\today}'
output:
  bookdown::pdf_document2:
    fig_caption: yes
    latex_engine: pdflatex
    template: tex-ms.tex
  pdf_document: default
  word_document: default
author:
- name: "Lucas Avezum"
  affiliation: "CentER and EBC, Tilburg University"
biblio-style: apsr
bibliography: ../../Auxiliar/mybibfile.bib
citecolor: black
csl: ../../Auxiliar/cambridge-journal-of-economics.csl
#anonymous: no
fontsize: 11pt
geometry: margin= 1in
header-includes: \usepackage{rotating, graphicx, dcolumn, longtable, float, siunitx, tabularx}
keywords: Internal ratings-based models, Risk-weighted assets dispersion, Procyclicality
linkcolor: black
abstract: The Basel II accord introduced a concave mapping from probabilities of default
  (PD) to capital requirements for banks adopting the internal ratings-based (IRB)
  approach. Therefore, total capital requirements depends not only on the portfolio's
  average PD but also on the shape of its PD distribution. Using a hand-collected
  dataset, I propose a method that decomposes the contribution of the PD distribution
  to capital requirements into an average component and a shape component. While a
  larger average component increases capital charges, a larger shape component generates
  capital savings. I investigate the relationship of these capital savings with the business cycle and find it to   be countercyclical. The results suggest that a recalibration of the current IRB framework can reduce considerably the procyclicality of capital regulation.
spacing: double
subtitle: ''
thanks: Preliminary draft.
JEL: G21, G28, G11
urlcolor: black
toc: FALSE
---

@@

library(tidyverse)            ## Data manipulation, pipe operator
library(data.table)           ## Command to bind lists
library(ggridges)             ## Density ridges plot
library(stargazer)            ## Latex tables
library(lfe)                  ## High-dimensional fixed effects
library(gridExtra)            ## Create grid for plots and tables
library(cowplot)              ## Create grid for plots and tables
library(DescTools)            ## Command to winsorize variables
library(ggpubr)               ## More plots
library(splitstackshape)      ## Command to expand data
library(latex2exp)            ## Use latex code in plots
library(lemon)                ## Plot brackets
#library(textablr)
#library(sandwich)
#library(lmtest)
#library(relaimpo)            ## Calculate relative importance measures

@@

\section{Introduction} \label{sec:intro}

The regulation on bank capital requirements has evolved substantially since the first international standard, the 1988 Basel Accord. At the heart of these changes is the link of capital charges to asset risk [@basel2004international; @BCBS2017basel; @EBA2019policy]. Risk sensitivity of capital requirements was present in the first accord and it was greatly enhanced in the Basel II framework by giving banks the option to calculate capital requirements using risk estimates obtained from their own models, known as the internal ratings-based (IRB) approach. The rationale for risk-linked capital charges is that regulating capital via a simple capital to asset ratio can incentivize banks to take on excessive risk [@koehn1980regulation; @kim1988risk; @gordy2010risk]. The rationale for allowing internal models is the better alignment of risk estimates to actual portfolio risk [@barakova2014banks] and the promotion of better risk management practices [@cucinelli2018credit].

However, both risk sensitivity and the use of internal models may have adverse consequences. First, linking capital charges to asset risks exacerbate the procyclicality of lending [@danielsson2001academic; @kashyap2004cyclical; @repullo2012procyclical; @behn2016procyclical]. As the overall credit quality deteriorates during economic downturns, capital requirements tighten, forcing banks to cut down lending. Second, the inherent flexibility of the IRB approach may allow for differences in capital requirements that do not reflect portfolio risk but rather modelling choices [@le2012revisiting; @mariathasan2014manipulation; @behn2016limits;@berg2017analysis; @ferribank].

@@

With this novel dataset, I start by documenting the relative importance of *Capital savings* to banks' overall RW. By extending the decomposition method proposed by @cannata2012inside, I find that *Capital savings* accounts for 27% of the RW variation across banks. I also find that *Capital savings* measures are negatively correlated to other RW components. This correlation implies that banks with higher average portfolio riskiness benefit the most from *Capital savings*. Lastly, although banks benefited from higher *Capital savings* as they moved their portfolio from the standardized to the IRB approach (known as the roll-out effect), the intensity of *Capital savings* has, on average, decreased in time. That is, for the same average PD, capital requirements have increased from 2008 to 2018.

@@

I further analyse the relationship of the PD distribution moments with GDP growth and find that during economic expansions average PD and PD variance decrease while the PD kurtosis and PD skewness increase. This evidence implies that distributions become more symmetric and with less extreme values during recessions, suggesting a flight to safety behavior by banks. As the recession reduces their existing portfolio credit quality (evidenced by the negative coefficient of average PD) banks respond by reducing risk-taking.

@@

The remainder of this paper is organized as follows. Section 2 describes the capital requirement formula, the *Capital savings* measure, and the hypothesis concerning this measure. Section 3 describes the collected data and analyses the relative importance of *Capital savings* in explaining differences in risk-weights across banks and time. Section 4 presents evidence on the countercyclical feature of *Capital savings*. Finally, section 5 concludes.  
\section{Institutional Background and Hypotheses} \label{sec:background}

\subsection{Capital Requirements under the IRB approach} \label{sec:cap_req}

In June 2004, the Basel Committee issued a revised framework on international convergence of capital measurements and capital standards [@basel2004international], known as Basel II, which serves as the basis for national rulemaking and implementation processes. At the heart of the revision was the link of capital charges to asset risk. Since the first accord from 1988 (Basel I), minimum capital requirements are calculated as a percentage of risk-weighted assets. Hence, the approach to define risk-weights is of major importance. In Basel I, risk-weights were flat within portfolio categories. For instance, all sovereign exposures were weighted by 0% and all corporate exposures by 100%. However, regulating capital via a simple capital to asset ratio can incentivize banks to take on excessive risk as the cost in terms of capital requirements is the same within a portfolio category [@koehn1980regulation; @kim1988risk; @gordy2010risk]. Therefore, since Basel II, financial institutions may choose between two approaches to calculate capital requirements for credit risk; the standardized approach (essentially a slightly modified version of the first accord) and the IRB approach.

In the IRB approach, capital requirements are derived from risk-weight formulas that take as input banks' own estimates of credit risk parameters. The rationale for allowing internal models is the better alignment of risk estimates to actual portfolio risk [@barakova2014banks] and the promotion of better risk management practices [@cucinelli2018credit]. The basic regulatory formula derives from a model characterized by the property that the risk-weight of each asset depends only on the estimated credit risk parameters for this asset and not on the composition of the portfolio, i.e. the model is portfolio invariant. This feature leads to a bottom-up approach where capital requirements are determined on the asset level and total requirement is simply the sum of asset individual requirement. Equation \ref{eq:RW_function} is the risk-weight formula at its most simple form:

\begin{equation}
\label{eq:RW_function}
\text{RW}(\text{LGD}_i, \text{PD}_i)= 12.5\text{LGD}_i\bigg[ N\bigg(\sqrt{\frac{1}{1-R}}N^{-1}(\text{PD}_i)+\sqrt{\frac{R}{1-R}}N^{-1}(0.999)\bigg)-\text{PD}_i\bigg],
\end{equation}
where $i$ is an asset in the portfolio and $N$ is the standard normal cumulative distribution. Basel accords segment credit portfolios into several categories (for instance, corporate, sovereign or retail), each having different additional features (maturity adjustments, risk-dependent correlation coefficient, among others) while keeping the basic structure of equation \ref{eq:RW_function}. Because of the portfolio invariant feature of the model, average risk-weight $\text{RW}^f$ is the exposure weighted average of all individual assets' risk-weights, as shown in equation \ref{eq:RW_factual}:

\begin{equation}
\label{eq:RW_factual}
\text{RW}^f= \frac{\sum_{i}\text{EAD}_i \text{RW}(\text{LGD}_i, \text{PD}_i)}{\sum_{i}\text{EAD}_i}= \sum_{i}\text{q}_i \text{RW}(\text{LGD}_i, \text{PD}_i).
\end{equation}
where $q_i= EAD_i/\sum_jEAD_j$. Thus, to calculated their capital requirements banks estimate three credit risk parameters: the exposure at default (EAD), the loss given default (LGD), and the probability of default (PD).[^1] The total amount of RWA for this portfolio is $\text{RW}^f\times \sum_i\text{EAD}_i$ and total capital requirements $\text{RWA}\times 8\%$.

[^1]: For wholesale exposures (corporate, sovereign and banks) the effective maturity of the asset is another parameter and for SMEs annual sales turnover is used to adjust the correlation coefficient.

\subsection{Probability of Default Distribution and Capital Savings} \label{sec:PD_capital_savings}

Figure \ref{fig:pdmap} plots the mapping from PD to RW according to equation \ref{eq:RW_function}. For this paper, the important characteristic of this mapping is that it is a concave function. The concavity of the PD-RW mapping together with the bottom-up method to aggregate capital requirements entail that actual capital requirements are lower than if they were calculated using the average PD to the entire portfolio. For clarity, consider two portfolios. The first contains a low PD asset and a high PD asset in proportions $q$ and $1-q$. The second has the same total amount as the first but contains only a mean PD asset, where mean PD is the mean of low PD and high PD weighted by their respective shares. Because of the concavity of the IRB formula and Jensen’s inequality, the capital requirement for the second portfolio is greater than the first. Figure \ref{fig:pdmap} illustrates the comparison. $\text{RW}^f$ and $\text{RW}^c$ are the average RW for the first portfolio and the second portfolio, respectively. More generally, total capital requirements do not depend only on the portfolio average PD but also on the shape of the PD distribution.

I propose a method that decomposes the contribution of the PD distribution to capital requirements into the average component and the shape component. The method is a generalization of the example in Figure \ref{fig:pdmap}: for each bank-year combination, I compare the actual capital requirement ($\text{RW}^f$) to the counterfactual case where the requirement for the entire portfolio is calculated using the portfolio’s average PD. The counterfactual RW is the average component of the PD distribution to capital requirements:

\begin{equation}
\label{eq:RW_counterfactual}
\text{RW}^c= \sum_{i}\text{q}_i \text{RW}(\text{LGD}_i,\overline{\text{PD}}),
\end{equation}
where the PD parameter for all assets is replaced by the portfolio weighted average. The latter is defined as the following:

@@

The difference between $\text{RW}^c$ and $\text{RW}^f$ measures any contribution to capital requirements from the PD distribution that is no related to the location of this distribution, i.e., it is a measure of the shape component. It can also be interpreted as the amount of *Capital savings* per unit of EAD due to the concavity of the IRB formula and the shape of the PD distribution. Equation \ref{eq:RW_savings} shows one of the possibilities to measure *Capital savings*:

\begin{equation}
\label{eq:RW_savings}
\text{RW}^{s}=\text{RW}^c-\text{RW}^f.
\end{equation}

Note that I defined *Capital savings* such that positive values translate to lower actual capital requirements and consequently $\text{RW}^s \geq 0$.

\subsection{Hypotheses on Capital Savings during the Business Cycle} \label{sec:hypotheses}

Overall, capital requirements under the IRB approach are procyclical. During economic downturns credit risk increases which raises capital requirements forcing banks to either raise equity or cut down lending. Theory and empirical evidence suggest that banks opt for the latter as raising equity is also more expensive during recessions [@danielsson2001academic; @kashyap2004cyclical; @repullo2012procyclical; @behn2016procyclical].

Between the two capital requirement components, the average component is a source of the procyclicality as long the relationship between the business cycle and PD is monotonic. Consider again the example shown in Figure \ref{fig:pdmap} with a portfolio containing a safe asset with PD equal to $PD_L$ and a risky asset with PD equal to $PD_H$ in proportions $q$ and $1-q$. In Appendix XX, I show that for $\text{PD} \in (0,0.25)$, $\text{RW}'>0$, $\text{RW}''<0$ and $\text{RW}'''>0$. I also show in Appendix XX that, on average, XX% of assets' PD lies within this range. For simplicity, I assume a constant $\text{LGD}_i=\text{LGD}$. So, the average component for this portfolio is:

\begin{equation}
\label{eq:RW_counter_example}
\text{RW}^{c}(q,\text{PD}_L, \text{PD}_H)=\text{RW}(q\text{PD}_L+(1-q)\text{PD}_H).
\end{equation}

Because $\text{RW}'>0$ and as long PD estimates are affected monotonic and negatively by GDP growth, the average component of capital requirements increases during economic downturns.

The direction of *Capital savings* during the business cycle is not trivial even when assuming a monotonic relationship between PD estimates and the business cycle. To explain the ambiguity, I consider three cases. But first, note that using the example in Figure \ref{fig:pdmap}, *Capital savings* can be written as the following:

\begin{equation}
\label{eq:RW_savings_example}
\text{RW}^{s}(q,\text{PD}_L, \text{PD}_H)=\text{RW}(q\text{PD}_L+(1-q)\text{PD}_H)-q\text{RW}(\text{PD}_L)+(1-q)\text{RW}(\text{PD}_H).
\end{equation}

We can find the effect of changes in the PD distribution to *Capital savings* by taking the total derivative of \ref{eq:RW_savings}:

\begin{equation}
\label{eq:RW_savings_diff}
\begin{aligned}
d\text{RW}^{s}(q,\text{PD}_L, \text{PD}_H)
&=\bigg[q\text{RW}'(q\text{PD}_L+(1-q)\text{PD}_H)-q\text{RW}'(\text{PD}_L)\bigg]d\text{PD}_L\\
&+\bigg[(1-q)\text{RW}'(q\text{PD}_L+(1-q)\text{PD}_H)-(1-q)\text{RW}'(\text{PD}_H)\bigg]d\text{PD}_H\\
&+\bigg[\text{RW}(\text{PD}_H)-\text{RW}(\text{PD}_L)-(\text{PD}_H-\text{PD}_L)\text{RW}'(q\text{PD}_L+(1-q)\text{PD}_H)\bigg]dq\\
\end{aligned}
\end{equation}

Next, consider the case that all assets are affected equally during the business cycles in term of PD, i.e.,  $d\text{PD}_H=d\text{PD}_L=d\text{PD}$. Also, consider for now no portfolio reallocation, i.e., $dq=0$. In this scenario equation \ref{eq:RW_savings_diff} becomes:

\begin{equation}
\label{eq:first_case}
\begin{aligned}
d\text{RW}^{s}
&=\bigg[\text{RW}'(q\text{PD}_L+(1-q)\text{PD}_H)-\bigg(q\text{RW}'(\text{PD}_L)+(1-q)\text{RW}'(\text{PD}_H)\bigg)\bigg]d\text{PD}.\\
\end{aligned}
\end{equation}

Because $\text{RW}'$ is a convex function and Jensen's inequality, if $d\text{PD}_H=d\text{PD}_H=d\text{PD}$ then $d\text{RW}^{s}/d\text{PD}<0$. In this case, *Capital savings* are procyclical, decreasing when PD estimates increase, i.e., during economic downturns. Figure \ref{fig:pdmapcasesym} shows graphically this first case.

Next, consider the case riskier assets are more affected during the business cycles in term of PD, i.e.,  $d\text{PD}_H>d\text{PD}_L$.For simplicity consider the extreme case of $d\text{PD}_H=d\text{PD}$ and $d\text{PD}_L=0$. In this scenario equation \ref{eq:RW_savings_diff} becomes:

\begin{equation}
\label{eq:second_case}
\begin{aligned}
d\text{RW}^{s}
&=(1-q)\bigg[\text{RW}'(q\text{PD}_L+(1-q)\text{PD}_H)-\text{RW}'(\text{PD}_H)\bigg]d\text{PD}\\
\end{aligned}
\end{equation}

Because $\text{RW}''<0$ and $\text{PD}_H>q\text{PD}_L+(1-q)\text{PD}_H$, if $d\text{PD}_H=d\text{PD}$ and $d\text{PD}$ then $d\text{RW}^{s}/d\text{PD}>0$. In this case, *Capital savings* are countercyclical, increasing when PD estimates increase. In other words, the marginal capital requirement decreases during economic downturns. Figure \ref{fig:pdmapcaseassym} shows graphically this second case.

Finally, consider the case where $d\text{PD}_H=d\text{PD}$, $d\text{RW}^{s}=0$ and $dq=0$. Using equation \ref{eq:RW_savings_diff}, we can find $d\text{PD}_L$ satisfying these conditions:

\begin{equation}
\label{eq:third_case}
\begin{aligned}
d\text{PD}_{L}
&=d\text{PD}\frac{(1-q)}{q}\frac{\bigg[\text{RW}'(\text{PD}_H)-\text{RW}'(q\text{PD}_L+(1-q)\text{PD}_H)\bigg]}{\bigg[\text{RW}'(q\text{PD}_L+(1-q)\text{PD}_H)-\text{RW}'(\text{PD}_L)\bigg]}\\
\end{aligned}
\end{equation}

Because $\text{RW}'<0$ and $\text{PD}_H>q\text{PD}_L+(1-q)\text{PD}_H>\text{PD}_L$, both denominator and numerator of equation \ref{eq:third_case} are negative and consequently, $d\text{PD}_L>0$. From the first case we also know that $d\text{PD}_L<d\text{PD}$. Hence, the difference between the impact of the business cycle on riskier and safer assets have to be sufficiently large such that *Capital savings* are countercyclical.

Whether *Capital savings* are pro- or countercyclical is an empirical question, as the direction of the cyclicality depends on how the economic cycle affects the shape of the PD distribution.

\section{Data and Capital Savings Measures} \label{sec:data}

The main measure of *Capital savings* used in this paper is the ratio of equation \ref{eq:RW_counterfactual} to equation \ref{eq:RW_factual} as follows:

@@

$\text{RW}^r$ is the rate at which a bank saves capital. For instance, a value of 1.5 means this portfolio would require 50% more capital in the absence of *Capital savings* to keep the same capital ratio. A value of one means that there are no *Capital savings* and implies a portfolio with a degenerate PD distribution. Any PD distribution with positive variance results in positive *Capital savings*.

@@

In the analysis I also consider the measure used in section \ref{sec:hypotheses} ($\text{RW}^s$), which is equivalent to dividing equation \ref{eq:measure2} by total EAD and therefore it is another measure of the intensity of *Capital savings*. Lastly, I calculate the Gini coefficient of the PD distribution as a robustness measure to my approach. According to the analysis shown in section \ref{sec:hypotheses} PD distributions with larger Gini coefficients should also generate larger *Capital savings*.

@@

Figure \ref{fig:savingsbank} shows that the *Capital savings* calculated with the compiled dataset using the main measure \ref{eq:measure1} (minus one to improve readability). Each point is a bank-year observations and the bars indicates the average value across time for each bank. There is substantial variation across banks with Goldman Sachs saving the highest at an average rate of 120% and Mediobanca the lowest at an average rate of 18.4%. Variation across time is considerable with, for instance, *Capital savings ratio* for Deutsche Bank going from 108.7% in 2010 to 61.4% in 2014. *Capital savings* also varies between and within countries. Figure \ref{fig:savingscountry}  aggregates the dataset by country. In the figure, each point is a bank observation and the bars are the average of these points. Variation between countries reaches its maximum at 50% (Denmark versus Norway) and within at 75% (between BNY Mellon and Goldman Sachs in the United States).

Besides the business cycle, I also test if bank-specific characteristics are associated to *Capital savings*. Several works relate bank-specific characteristics to either total RWA or average RW [@vallascas2013risk; @mariathasan2014manipulation; @ferribank; @gropp2018banks]. Table \ref{tab:summary} shows summary statistics for the all the dependent variables, the main variable of interest (GDP growth rate), and all the other variables used in the analysis.

The first set of variables captures banks’ opportunities and incentives to affect RWA by means of capital arbitrage. First, I control for bank size using the log of total assets (in millions of US dollars). The expectation regarding the impact of bank size is ambiguous. On the one hand, larger banks might be able to adopt more sophisticated internal risk models and better exploit modelling choices to increase capital savings. On the other hand, larger banks might be subject to closer regulatory scrutiny and, as a result, may find it more difficult to manipulate their risk estimates. I also include the regulatory capital ratio as control. Better capitalized banks could be subject to less regulatory scrutiny [@calem1999impact] and, consequently, in a better position to engage in regulatory arbitrage. However, the incentives to manipulate risk parameters should be lower for banks with high capital ratios as they are at a greater distance to the minimum requirement. Thus, the expected effect of capital ratio on capital savings is also ambiguous. Next, I add banks' profitability using return on asset (ROA) and return on equity (ROE) with both measured before taxes. Profitable banks face fewer incentives to understate the riskiness of their assets and to engage in regulatory arbitrage more generally. Finally, I study the relationship of *Capital savings* to risk-taking. I add the ratio of non-performing loans to total loans to test this proposition.

\section{Capital Savings and the Business Cycle} \label{sec:cycle}

In this section I investigate the relationship between *Capital savings* and the business cycle. I start by explaining the theoretical ambiguous relationship between *Capital savings* and the business cycle. Next I show that *Capital savings* are countercyclical, i.e., given an average portfolio PD, banks need more capital in good times than in bad times to reach the same capital ratio. Finally, I provide evidence that this relationship is not purely mechanical but depends on how banks reallocate their assets through the business cycle.

@@

The dependent variable is the yearly change in a variable associated to portfolio $j$ for bank $i$ at year $t$. The change in the logarithm of real GDP per capita used comes from the country where the bank has its headquarter. All regressions include bank and year fixed effects and robust standard errors adjusted for clustering at the country level are reported in parentheses.

Column 1 Table \ref{tab:main} shows results using equation \ref{eq:measure1}, my preferred measure of *Capital savings*.  shows that total *Capital savings* are countercyclical. According to regression 1 an increase of 1.6 percentage points in GDP growth rate (one standard deviation) decreases total *Capital savings* growth rate by 0.7 percentage points on average. The effect is economically sizable representing 14% of total *Capital savings* standard deviation in the sample.

@@

\subsection{Selection and Reallocation Across Portfolios} \label{sec:cross_reallocation}
